Numerical analysis of the linear vertical shear instability in astrophysical disks

### 2014-12-23 linear_adia.tex
added numerical results for variable gmma

\subsubsection{Comparison with $\gamma=1$}
We can compare growth rates obtained for $\gamma=2$ to those for
$\gamma=1$ in the previous section. If we compare growth rates at their 
respective optimum radial wavelengths, we find
that for $\gamma=2$ the \emph{maximum} growth rate, occuring
at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
rate for $\gamma=1$. That is,
\begin{align}
  &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
      q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
  &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
           & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
\end{align}

@@

\subsubsection{Qualitative expectations in the presence of thermal
  relaxation}
We have shown that when $\gamma>1$ so that there is a positive
vertical entropy gradient and perturbations evolve
adiabatically ($t_c\to\infty$) that the VSI is strongly
surpressed. This is the limit where the term proportional to
$t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
becomes negligible compared to the term proportional to
$\gamma/\Gamma-1$. The latter term is associated with the vertical
entropy gradient, or buoyancy frequency squared, $N_z^2$.

>>>>

\subsubsection{Comparison with $\gamma=1$}\label{adia_compare_gam1}
We can compare growth rates obtained for $\gamma=2$ to those for
$\gamma=1$ in the previous section. If we compare growth rates at their 
respective optimum radial wavelengths, we find
that for $\gamma=2$ the \emph{maximum} growth rate, occuring
at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
rate for $\gamma=1$. That is,
\begin{align}
  &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
      q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
  &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
           & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
\end{align}

@@

\subsubsection{Qualitative condition for buoyancy to be negligible in
  the presence of thermal relaxation}
We have shown that when $\gamma>1$ so that there is a positive
vertical entropy gradient and perturbations evolve
adiabatically ($t_c\to\infty$) that the VSI is strongly
surpressed. This is the limit where the term proportional to
$t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
becomes negligible compared to the term proportional to
$\gamma/\Gamma-1$. The latter term is associated with the vertical
entropy gradient, or buoyancy frequency squared, $N_z^2$.

### 2014-12-23 linear_iso.tex
added numerical results for variable gmma

\subsection{Nearly-isothermal perturbations}
We can use the explicit solution developed above to examine the effect
of cooling $t_c$ on perturbations with $\gamma>1$. We shall consider
the limit $t_c\to0$, for which perturbations can be considered
nearly-isothermal.

@@

Since we have developed solutions for $\beta\equiv 0$, we may ask how
do the eigenfrequencies and eigenfunctions change when we change
$\beta$ from zero to a small but finite value? For sufficiently small
$\beta$, we expect the new solutions to only differ slighly from that
above. Then we may linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} to
determine the effect of adding small finite cooling.

Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, W\to W+\delta W,\, Q \to Q+\delta
  Q,\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align} 
where $W=Q$ is the polynomial solution described in \S\ref{iso_poly}
and $\hat{\sigma}^2$ is given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso1}---\ref{nearly_iso2} and keeping only first order
terms, we obtain
\begin{align}
  &2\hat{\sigma}\delta\hat{\sigma}\left(1+\hat{k}^2\right)W +
  \hat{\sigma}^2 \left(\delta Q + \hat{k}^2\delta W\right)\notag\\
  &= \zhat \left(1+\ii\epsilon q \hat{k}\right)\left[\delta W^\prime -
    \zhat \left(\delta W - 
      \delta Q\right)\right] - \delta W^{\prime\prime}\notag\\
  &\phantom{=} + \left[\zhat \left(\delta W -
     \delta Q\right)\right]^\prime,\label{nearly_iso3}\\
&\ii\hat{\sigma} \left(\delta W - \delta Q\right) =
\delta\beta\left(\gamma-1\right)\left(\hat{\sigma}^2W - \hat{z}W^\prime\right).\label{nearly_iso4}
\end{align}

We could proceed further by considering the set of polynomial
solutions for $W$ described in \S\ref{iso_poly} altogether, but it is
simplest to look at the fundamental mode. Thus
we consider the `basic state'
\begin{align*}
  W = \hat{z},\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
  \ii\epsilon q \hat{k},
\end{align*}
then we can assume
\begin{align}
  \delta W = b \zhat^3,
\end{align}
where $b$ is a constant to be determined. Eliminating $\delta Q$ from
Eq. \ref{nearly_iso3}---\ref{nearly_iso4}, inserting the above
expressions for $W$ and $\delta W$ and balancing coefficients of
$\zhat$ and $\zhat^3$, we find
\begin{align}
 & 2\left(1+\ii\epsilon q \hat{k}\right) \delta\hat{\sigma} \notag\\& = -6b\hat{\sigma} -
2\ii \delta \beta\left(\gamma-1\right)\left(\hat{\sigma}^2-1\right) - \ii
  \hat{\sigma}^2\delta\beta\left(\gamma-1\right),\\
& 0=  2b\hat{\sigma}+\ii\delta\beta
\left(\gamma-1\right)\left(\hat{\sigma}^2 - 1\right),
\end{align}
which gives
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
    q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
   \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
        \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\nu < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by finite cooling is ineffective at very large or very small scales.

>>>>

\subsection{Nearly-isothermal perturbations}
The explicit solution developed in \S\ref{iso_poly} is valid for
non-isothermal perturbations $\gamma>1$ if the thermal relaxation
(cooling) timescale $t_c\equiv0$. We may ask how
do the eigenfrequencies and eigenfunctions change when we change
$t_c$ from zero to a small but finite value? For sufficiently small
$t_c$, for which the perturbations may be considered
nearly-isothermal, we expect the new solutions to only differ slighly
from that above.

% Then we may linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} to
% determine the effect of adding small finite cooling.

@@

We linearize Eq. \ref{nearly_iso1}---\ref{nearly_iso2} about a known
solution for $\beta\equiv0$. Let 
\begin{align}\label{nearly_iso_pert}
  \beta \to 0 + \delta\beta,\, W\to W+\delta W,\, Q \to Q+\delta
  Q,\,\hat{\sigma} \to \hat{\sigma} + \delta\hat{\sigma}, 
\end{align} 
where $W=Q$ is the polynomial solution described in \S\ref{iso_poly}
with $\hat{\sigma}^2$ given by Eq. \ref{sig2_iso}. Inserting
Eq. \ref{nearly_iso_pert} into
Eq. \ref{nearly_iso1}---\ref{nearly_iso2} and keeping only first order
terms, we obtain
\begin{align}
  &2\hat{\sigma}\delta\hat{\sigma}\left(1+\hat{k}^2\right)W +
  \hat{\sigma}^2 \left(\delta Q + \hat{k}^2\delta W\right)\notag\\
  &= \zhat \left(1+\ii\epsilon q \hat{k}\right)\left[\delta W^\prime -
    \zhat \left(\delta W - 
      \delta Q\right)\right] - \delta W^{\prime\prime}\notag\\
  &\phantom{=} + \left[\zhat \left(\delta W -
     \delta Q\right)\right]^\prime,\label{nearly_iso3}\\
&\ii\hat{\sigma} \left(\delta W - \delta Q\right) =
\delta\beta\left(\gamma-1\right)\left(\hat{\sigma}^2W - \hat{z}W^\prime\right).\label{nearly_iso4}
\end{align}
This is a pair of inhomogeneous ODEs with a forcing by $W$. The
homogeneous problem is, of course, just the ODE for $\beta=0$ solved
in \S\ref{iso_poly}. Hence we only seek the particular solution.

We could proceed by considering the entire set of polynomial
solutions for $W$, but it is simplest to look at the fundamental
mode. Thus we consider the `basic state' 
\begin{align*}
  W = \hat{z},\quad \hat{\sigma}^2\left(1+\hat{k}^2\right) = 1 +
  \ii\epsilon q \hat{k},
\end{align*}
then
\begin{align}
  \delta W = b \zhat^3,
\end{align}
where $b$ is a constant to be determined. Eliminating $\delta Q$ from
Eq. \ref{nearly_iso3}---\ref{nearly_iso4}, inserting the above
expressions for $W$ and $\delta W$ and balancing coefficients of
$\zhat$ and $\zhat^3$, we find
\begin{align}
 & 2\left(1+\ii\epsilon q \hat{k}\right) \delta\hat{\sigma} \notag\\& = -6b\hat{\sigma} -
2\ii \delta \beta\left(\gamma-1\right)\left(\hat{\sigma}^2-1\right) - \ii
  \hat{\sigma}^2\delta\beta\left(\gamma-1\right),\\
& 0=  2b\hat{\sigma}+\ii\delta\beta
\left(\gamma-1\right)\left(\hat{\sigma}^2 - 1\right),
\end{align}
where the expression for $\hat{\sigma}^2$ was used. Solving for
$\delta\hat{\sigma}$ gives 
\begin{align}\label{nearly_iso_dsig}
  \delta \hat{\sigma} =
  -\frac{\ii\left(\gamma-1\right)\hat{k}^2\left(\ii\epsilon
    q - \hat{k}\right)^2}{2\left(1+\ii\epsilon q \hat{k}\right)\left(1+\hat{k}^2\right)^2}\delta\beta.
\end{align}
The imaginary part of Eq. \ref{nearly_iso_dsig} is
\begin{align}
   \delta\hat{\nu} =
  -\frac{\left(\gamma-1\right)\hat{k}^2 \left(\hat{k}^2 -
      2\epsilon^2q^2\hat{k}^2 - \epsilon^2q^2\right)}{2\left(1+\epsilon^2 q^2
        \hat{k}^2\right)\left(1+\hat{k}^2\right)^2}\delta\beta.  
\end{align}
Since $\epsilon \ll 1$, introducing finite cooling $\delta\beta>0$
implies $\delta\hat{\nu} < 0$, i.e. stabilization. Interestingly, for both
$\hat{k}\to0$ and $\hat{k}\to\infty$ we find $\delta\hat{\nu}\to0$, so
stabilization by finite cooling is ineffective at very large or very
small scales. We plot Eq. \ref{nearly_iso_dsig} in Fig. \ref{domegadbeta} with
$\epsilon=0.05$, $q=-1.0$ and $\gamma=1.4$. For this disk model the
optimum wavenumber for stabilization is $\hat{k}\sim 6$.

% This contrasts to the case without vertical shear
% ($q=0$), in which $\p\hat{\sigma}/\p\beta\to\mathrm{constant}$ as 
% $\hat{k}\to\infty$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/domegadbeta}
  \caption{Rate of change of the fundamental VSI eigenfrequency
    $\sigma$ with respect to the introduction of a small but finite
    cooling time. The plotted quantity is $\p\hat{\sigma}/\p\beta$ at $\beta=0$, evaluated using
    Eq. \ref{nearly_iso_dsig} with 
    disk parameters $\epsilon=0.05$, $q=-1$ and $\gamma=1.4$. The real
    (imaginary) part of $\p\hat{\sigma}/\p\beta$ is shown as the solid
    (dotted) lines. 
    \label{domegadbeta}}  
\end{figure}

One effect of introducing finite thermal relaxation is buoyancy. We
show in the next section that buoyancy strongly stabilizes the
VSI. The solution $\delta W\propto\zhat^3$ then makes sense because it
is most significant for large $\zhat$, i.e. away from the midplane
where the effect of bouyancy first appears as $t_c$ is increased.

### 2014-12-23 numerical.tex
added numerical results for variable gmma

\subsubsection{Other adiabatic indices}

\subsection{Effect of thermal relaxation}
{\bf vertical isothermal disk with tcool}

\subsection{Vertically non-isothermal disks}
{\bf big gamma larger than 1 but equal to small gamma, with tcool}

\subsection{Non-uniform thermal relaxation} 
{\bf variable tcool, large at midplane, small in atmosphere. simplest
  model is to multiply tcool by background density}

>>>>

\subsubsection{Other adiabatic indices}
In Fig. \ref{adia_growth_num} we plot the eigenfrequencies of the
fundamental VSI mode for a range of $\gamma$ values. This plot is to
be compared with Fig. \ref{adia_growth}.

% The numerically-determined eigenfrequencies are qualitatively
% consistent with the discussion in \S\ref{adia_compare_gam1}.
There is good agreement between Fig. \ref{adia_growth_num} and
Fig. \ref{adia_growth} for $k_xH_\mathrm{iso}\gtrsim10$. There is a
mistach for smaller wavenumbers, especially in the growth
rates. Actual growth rates decrease slower as $k_xH_\mathrm{iso}\to0$
than that in Fig. \ref{adia_growth}. Disagreement at small
$k_xH_\mathrm{iso}$ is not surprising because the low-frequency
approximation used to obtain Fig. \ref{adia_growth} becomes invalid since
$|\omega|\sim\Omega_k$ for small $k_x$.

Given the number of approximations made in \S\ref{adia_compare_gam1},
Fig. \ref{adia_growth_num} compares well with
Fig. \ref{adia_growth}. The main qualitative features of the full
numerical solution in Fig. \ref{adia_growth_num} are captured by
Fig. \ref{adia_growth}: stabilization for
$k_xH_\mathrm{iso}\to0$ and $k_xH_\mathrm{iso}\to\infty$,
stabilization for increasing $\gamma$ (but ineffective at small $k_x$), 
and the shift of the most unstable wavenumber to smaller values as
$\gamma$ is increased.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0cm]{figures/compare_eigen_real1}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 1cm]{figures/compare_eigen_imag1}
  \caption{Real frequency (top) and growth rate 
    (bottom) of the fundamental VSI mode in nearly-vertically   
    isothermal disks ($\Gamma=1.011$) with $\epsilon=0.1$ and  
    $q=-1$, subject to adiabatic perturbations
    ($\beta=10^9$). The eigenfrequencies are calculated by numerically
    solving the full linear eigenvalue problem. \label{adia_growth_num}}   
\end{figure}

\subsection{Effect of thermal relaxation}
We now introduce thermal relaxation. Here we consider a slightly
different disk model with $\epsilon=0.05$ and
$\gamma=1.4$. This model was used by \citep{nelson13}, where the
authors find a cooling timescale of $\beta\geq 0.1$ was sufficient to
stabilize the VSI in non-linear hydrodynamic simulations.

% \subsection{Vertically non-isothermal disks}
% {\bf big gamma larger than 1 but equal to small gamma, with tcool}

% \subsection{Non-uniform thermal relaxation} 
% {\bf variable tcool, large at midplane, small in atmosphere. simplest
%   model is to multiply tcool by background density}
